Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-08-22 analysis/paper/paper.Rmd
replace culture contact with presence, arrival, or interaction, change the font color of Taiwan map

---
title: "Ornaments as indicators of social change before and after European contact at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  In East Asia, direct European colonial rule on local communities was much less pronounced than in other parts of the world. In this region, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Using trade ornaments recovered from Kiwulan (1400-1900 AD) in northeastern Taiwan, I explore the indirect impacts of European colonial activities on local societies. I examine the cultural contexts, variety, and spatial distributions of trade ornaments in three time episodes relating different scales of culture contact. The results indicate the diversity of ornaments was greater during the European contact period compared to previous periods and their spatial distribution was clustered. This hints at increasing social inequality in a colonial context. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European contact.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less \pronounced than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Studying the consumption of foreign goods can improve our understanding of the negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Rubertone2000; @Scaramelli2005; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005].

@@

Although mountains isolated Yilan by land, rivers and seas provided a way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan, and Han Chinese. The indigenous people of Yilan established their own exchange network that has been described as part of an "inter-insular trade" system [@Chen2005, p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis before European contact. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006]. Trading activities brought a wide variety of objects into indigenous communities in Yilan. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore archaeological evidence from Yilan spanning from the late Iron Age to the historical period to address the question of whether indirect colonial influences on the indigenous populations can be detected. I focus on the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf. @Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine the potential of ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new patterns in the spatial distributions of ornaments at Kiwulan that indicate increasing social inequality. I hypothesize that after 1662 when European contact declined and Chinese contact became the dominant international tradition partner, there was a decline in the use and discard of ornaments at Kiwulan due to overall population declines and acculturation in the Chinese community. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European contact, and a lesser diversity after Chinese contact, I compare frequencies of ornament types between different time periods. To test the prediction that after European contact were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

@@

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. Despite most Kavalan people today not knowing the specific origin of their agate beads, the high value of beads is still indicated in oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to the European contact period, there are fewer mentions of beads in Chinese contact period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

@@

```{r tidy-data, eval = TRUE}
# summarise the ornaments by layers
# L5 and L6 represent pre-European contact period, 
# L4 was the time of European contact, and 
# L2 and L1 was the period of Chinese contact. 
ornaments_period <- 
kwl_upper %>% 
  filter(!is.na(`6-layers`)) %>% 
  mutate(period = case_when(
    `6-layers` %in% 1:2 ~ "Chinese Contact",
    `6-layers` == 4 ~ "European Contact",
    `6-layers` %in% 5:6 ~ "Before European Contact",
     TRUE ~ "other"
  )) %>% 
  filter(period != "other") %>% 
  mutate(period = factor(period,
                       level = c("Before European Contact", 
                                 "European Contact", 
                                 "Chinese Contact"))) %>% 
  mutate(Length = as.numeric(`Length(mm)`),
         Thick = as.numeric(`Thick(mm)`),
         Width = as.numeric(`Width(mm)`),
         Perforation1 = as.numeric(`Perforation1(mm)`)) %>% 
  rowwise() %>% 
  mutate(Perforation_average =
           mean(c(`Perforation1`,`Perforation2(mm)`), 
                na.rm = TRUE))
```

@@

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700 - 1200 AD) and a Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age to the historical period starting from the 17th century with European contact in Taiwan.

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5 and L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. In the archaeological record, the approximate indicators of the start of European contact at Kiwulan are the appearance of An-ping jars and stonewares (martavans/martaban) that were largely introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods.

European colonization ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese contact period at Kiwulan is the large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration to Yilan can be seen from official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Contact`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artifact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)
```

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$) shows that there is a non-random difference between the three time periods. It indicates that most ornament types were present before European contact. Ornament frequencies reached a peak during European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European contact period and a further decrease in the Chinese contact period.

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across contact periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.png"),
       width = 190, 
       height = 150,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.png"))
```

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture contact instead of effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the Europeans. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before European contact, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European contact period, ornaments were more widespread, with some clusters on the northern part. During the Chinese contact period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European contact, such as golden beads and agate beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, the cluster of golden beads was found at the northern part, while the cluster of agate beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and seem randomly distributed across the area. In the Chinese Contact period, both the amount and density of different classes of ornaments decreased.

```{r point-pattern-data-tidy}
# get centroid and make a list of three periods
ornaments_period_join_cen_all <-
  ornaments_period_join %>% 
  mutate(lon = jitter(map_dbl(geometry, ~st_point_on_surface(.x)[[1]])),
         lat = jitter(map_dbl(geometry, ~st_point_on_surface(.x)[[2]]))) 
ornaments_period_join_cen <- 
  ornaments_period_join_cen_all %>% 
  group_by(period) %>% 
  group_split() %>% 
  set_names(c("Before European Contact",
              "European Contact",
              "Chinese Contact"))
```

@@

I used point pattern analysis to examine whether the distribution of ornaments was uneven in non-random ways that could reflect changes in social inequality. Point pattern analysis is a way to assess whether the distribution of sites or artifacts represent some “hotspots” produced by non-random processes, implying the influence of social processes [@Bevan2016; @Ducke2015]. Hotspots in the pattern of ornaments could indicate changes in social organization due to culture contact. To explore ornament distributions I used the intensity approach that focuses on the average density of points across space. During excavation, ornament locations were only recorded to the square (4 x 4 m). To prepare the location data for point pattern analysis, I assigned each ornament to a random coordinate pair located in square it was recovered from. The next step was to subset the ornaments into three groups for three time periods. Finally I computed the kernel densities for each time period for comparison.

@@

The results show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period (Figure \@ref(fig:plot-all-densities)). Those maps present three consistent sub-regions with a core area that shifts over time. The distribution might indicate different social groups who possessed more ornaments. The multiple groups during European contact period might reflect more unequal consumption of ornaments across the site, relative to other periods. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate the pattern, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.

@@

To test this hypothesis about the randomness of spatial locations, I used a Monte Carlo method to simulate average nearest-neighbor distances (ANN). Our null hypothesis is that the distribution of ornaments is consistent with a completely random process. We simulated the locations of ornaments across the space 1000 times for each time period. The observed data was compared to the random patterns generated by simulated processes to determine if our data are randomly distributed or not. Figure \@ref(fig:hypo-test-distribution) shows the distributions of the ANN distances calculated on 1000 simulations of ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European contact period, which means the ornaments have non-random clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese contact period, with about one third of the simulated values are greater than our observed ANN value. This testing reveals that the clustered distributions of ornaments during the pre-European and European contact periods might reflect the presence of different social groups. Moreover, the clustering of ornaments during the European contact period is highly non-random, potentially indicating that some of the social changes leading to a concentration of power to control the distribution of ornaments at Kiwulan.

@@

# Acknowledgements

I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the ornaments used in this study. I thank Shui-jin Chiu, Shu-hui Jian, Jiun-yao  Lai, and the staff in archaeology lab for their invaluable assistance in preparing samples for analysis. I also thank Yu-pei Chen for the discussion regarding the site at the earliest stage of fieldwork. This work was supported in part by the Fritz Graduate Fellowship and travel grants from the Department of Anthropology at University of Washington. I especially thank Ben Marwick for providing academic support during this project, helping me with polishing figures with R langauge, and providing feedback on early drafts of this manuscript.

>>>>

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  In East Asia, direct European colonial rule on local communities was much less pronounced than in other parts of the world. In this region, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Using trade ornaments recovered from Kiwulan (1400-1900 AD) in northeastern Taiwan, I explore the indirect impacts of European colonial activities on local societies. I examine the cultural contexts, variety, and spatial distributions of trade ornaments in three periods relating to different cultural interactions. The results indicate the diversity of ornaments was greater during the European presence period compared to previous periods and their spatial distribution was clustered. This hints at increasing social inequality in a colonial context. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Studying the consumption of foreign goods can improve our understanding of the negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Rubertone2000; @Scaramelli2005; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005].

@@

Although mountains isolated Yilan by land, rivers and seas provided a way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan, and Han Chinese. The indigenous people of Yilan established their own exchange network that has been described as part of an "inter-insular trade" system [@Chen2005, p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis before European were present. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006]. Trading activities brought a wide variety of objects into indigenous communities in Yilan. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore archaeological evidence from Yilan spanning from the late Iron Age to the historical period to address the question of whether indirect colonial influences on the indigenous populations can be detected. I focus on the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf. @Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine the potential of ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European arrival due to participation in a larger scale exchange network. I also predict that after the European presence became established there will be new patterns in the spatial distributions of ornaments at Kiwulan that indicate increasing social inequality. I hypothesize that after 1662 when the European presence declined and the Chinese became the dominant international tradition partner, there was a decline in the use and discard of ornaments at Kiwulan due to overall population declines and acculturation in the Chinese community. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European arrival, and a lesser diversity after the Chinese presence became established, I compare frequencies of ornament types between different time periods. To test the prediction that after European arrival were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

@@

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. Despite most Kavalan people today not knowing the specific origin of their agate beads, the high value of beads is still indicated in oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

@@

```{r tidy-data, eval = TRUE}
# summarise the ornaments by layers
# L5 and L6 represent pre-European contact period, 
# L4 was the time of European contact, and 
# L2 and L1 was the period of Chinese contact. 
ornaments_period <- 
kwl_upper %>% 
  filter(!is.na(`6-layers`)) %>% 
  mutate(period = case_when(
    `6-layers` %in% 1:2 ~ "Chinese Presence",
    `6-layers` == 4 ~ "European Presence",
    `6-layers` %in% 5:6 ~ "Before European Contact",
     TRUE ~ "other"
  )) %>% 
  filter(period != "other") %>% 
  mutate(period = factor(period,
                       level = c("Before European Contact", 
                                 "European Presence", 
                                 "Chinese Presence"))) %>% 
  mutate(Length = as.numeric(`Length(mm)`),
         Thick = as.numeric(`Thick(mm)`),
         Width = as.numeric(`Width(mm)`),
         Perforation1 = as.numeric(`Perforation1(mm)`)) %>% 
  rowwise() %>% 
  mutate(Perforation_average =
           mean(c(`Perforation1`,`Perforation2(mm)`), 
                na.rm = TRUE))
```

@@

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700 - 1200 AD) and a Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age to the historical period starting from the 17th century with the European presence in Taiwan.

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period (L5 and L6), European period (L4), and Chinese period (L2 and L1).

I identify the first European arrival at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. In the archaeological record, the approximate indicators of the start of the European influence at Kiwulan are the appearance of An-ping jars and stonewares (martavans/martaban) that were largely introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods.

European colonization ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration to Yilan can be seen from official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artifact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)
```

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$) shows that there is a non-random difference between the three time periods. It indicates that most ornament types were present before European contact. Ornament frequencies reached a peak during European period and then dropped during the Chinese period, especially for golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period.

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.png"),
       width = 190, 
       height = 150,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.png"))
```

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the Europeans. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before the European arrival, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European period, ornaments were more widespread, with some clusters on the northern part. During the Chinese period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European period, such as golden beads and agate beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, the cluster of golden beads was found at the northern part, while the cluster of agate beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and seem randomly distributed across the area. In the Chinese period, both the amount and density of different classes of ornaments decreased.

```{r point-pattern-data-tidy}
# get centroid and make a list of three periods
ornaments_period_join_cen_all <-
  ornaments_period_join %>% 
  mutate(lon = jitter(map_dbl(geometry, ~st_point_on_surface(.x)[[1]])),
         lat = jitter(map_dbl(geometry, ~st_point_on_surface(.x)[[2]]))) 
ornaments_period_join_cen <- 
  ornaments_period_join_cen_all %>% 
  group_by(period) %>% 
  group_split() %>% 
  set_names(c("Before European Contact",
              "European Presence",
              "Chinese Presence"))
```

@@

I used point pattern analysis to examine whether the distribution of ornaments was uneven in non-random ways that could reflect changes in social inequality. Point pattern analysis is a way to assess whether the distribution of sites or artifacts represent some “hotspots” produced by non-random processes, implying the influence of social processes [@Bevan2016; @Ducke2015]. Hotspots in the pattern of ornaments could indicate changes in social organization due to culture interaction. To explore ornament distributions I used the intensity approach that focuses on the average density of points across space. During excavation, ornament locations were only recorded to the square (4 x 4 m). To prepare the location data for point pattern analysis, I assigned each ornament to a random coordinate pair located in square it was recovered from. The next step was to subset the ornaments into three groups for three time periods. Finally I computed the kernel densities for each time period for comparison.

@@

The results show that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period (Figure \@ref(fig:plot-all-densities)). Those maps present three consistent sub-regions with a core area that shifts over time. The distribution might indicate different social groups who possessed more ornaments. The multiple groups during European period might reflect more unequal consumption of ornaments across the site, relative to other periods. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate the pattern, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.

@@

To test this hypothesis about the randomness of spatial locations, I used a Monte Carlo method to simulate average nearest-neighbor distances (ANN). Our null hypothesis is that the distribution of ornaments is consistent with a completely random process. We simulated the locations of ornaments across the space 1000 times for each time period. The observed data was compared to the random patterns generated by simulated processes to determine if our data are randomly distributed or not. Figure \@ref(fig:hypo-test-distribution) shows the distributions of the ANN distances calculated on 1000 simulations of ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-random clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. This testing reveals that the clustered distributions of ornaments during the pre-European and European periods might reflect the presence of different social groups. Moreover, the clustering of ornaments during the European period is highly non-random, potentially indicating that some of the social changes leading to a concentration of power to control the distribution of ornaments at Kiwulan.

@@

# Acknowledgments
